---
title: "Lab2"
author: "Glenn (Ted) Dunmire"
date: "November 13, 2015"
output: html_document
---

##Question 1: Concepts


_1.1 (1 points) Define the term stochastic process._

A stochastic process is defined as a statistical phenomenon evolving according to probabilistic laws. 

_1.2 (2 points) Define the term time series. What is the difference between a stochastic process and a time series?_

A time series is a set of observations generated sequentially in time. A time series arises as the result of a stochastic process. 

###EXPAND ON THIS

_1.3 (3 points) In your own words, discuss the mean function, variance function, and measure of dependency structure in the context of time series (week 8 - 10) and compare them with those we studied in classical linear model (week 1 - 7)._

Mean is largely meaningingless in a time series context
##TBD - STILL WORKING ON

_1.4 (2 points) Define strict and weak statonarity_

Strict stationarity is when the joint distributions for all time periods are the same. Formally it is if 
$$F(x_t1, . . ., x_tn)$$ and $$F(x_t1+m, . . ., x_tn+m)$$ are equal for all (t1, . . ., tn) and m. 

Weak stationarity is when a time series is stationary in mean and variance and if its autocovariance depends on time (k) and can be written as $$\gamma^{k}$$

_1.5 (3 points) Give an example of a time series in real life. Describe the series. Evaluate (not empirical work is needed) whether or not the series can be modeled using the class of autoregressive models?_

##NEED TO FIND ONE

_1.6 (4 points) In your own words, define and describe partial autocorrelation function (PACF). Why is it not enough just to autocorrelation functon (ACF) to capture the dependency of a series?_

##NEED TO DEFINE

##Question 2

_Determine if each of the following models is stationary_

Use the polyroot function and if the absolute value of all roots is > 1 then stationary

2.1 (5 points) $$zt = 0.95zt−1 + ωt$$

```{r}
polyroot(c(1, 0.95))
```

The absolute value is greater than 1, therefore this model is stationary. 

2.2 (5 points) $$zt = 0.8zt−1 + 0.3zt−2 + ωt$$

```{r}
polyroot(c(1, 0.8, 0.3))
```

Both coefficients are greater than 1 in absolute value, therefore this model is stationary. 

2.3 (5 points) $$zt = −0.5zt−1 + 0.5zt−2 + ωt$$

```{r}
polyroot(c(1, -0.5, 0.5))
```

##DON'T THINK SO?

2.4 (5 points) $$zt = zt−1 + 0.4zt−2 + ωt$$

```{r}
polyroot(c(1, 1, 0.4))
```

Both roots are greater than 1 in absolute value, therefore the model is stationary. 

2.5 (5 points) $$zt = −0.5zt−1 − 0.25zt−2 + ωt$$

```{r}
polyroot(c(1, -0.5, -0.25))
```

Both roots are greater than 1 in absolute value, therefore the model is stationary. 